GPU Efficient Texture Based Bezier Curve Evaluation

### 2015-07-19 Paper/paper.tex
Replaced order with Degree

\begin{abstract}
\small
Modern graphics techniques expose internal parameters to allow re-use and to help separate the technical implementation from artistic usage cases.  A popular choice is to expose parameters as customizable curves and to quantize the curves into textures.  Quantization leads to either lower quality results, or more texture memory being used to accommodate higher sampling frequencies.  The technique presented in this paper leverages the capabilities of GPU texture samplers to allow more efficient storage and evaluation of both integral and rational Bezier curves of any order, resulting in higher fidelity for similar costs.  Piecewise curves, B-Splines and NURBS are addressed, and there are also limited applications towards vector graphics.
\end{abstract}

@@

This paper shows a third method where:
\begin{itemize}
  \item The control points of the curves are encoded in a texture.
  \item The texture sampler calculates points on the curve before the data reaches the shader program.
  \item It gives accuracy results closer to that of shader constant curves, while having less calculation overhead.
  \item The technique can support both integral and rational Bezier curves of any order and can also be used for piecewise curves.
  \item The curve type must be decided on in advance, like when using the shader constant method.
  \item There are limited applications towards vector graphics.
\end{itemize}

@@

The core of this technique is that the $N$-dimensional linear texture interpolation capabilities on the GPU can be mathematically equivalent to De Casteljeau's algorithm for evaluating Bezier curves.  We can use linear texture sampling on textures of increasing dimensionalities to have the texture sampler calculate points on curves of increasing order.

@@

One dimensional linear texture sampling is just the linear interpolation between two data points.  This is equivalent to the De Casteljeau algorithm when evaluating a linear curve of order 1.  In both cases, it is just linear interpolation between two values.

@@

The De Casteljeau algorithm for evaluating a quadratic curve (order 2) is also just a linear interpolation between two linear interpolations.  If the control points of the quadratic curve are $P_0,P_1,P_2$, let's call the first interpolation $\overline{P_0P_1}$ which is the interpolation between $P_0$ and $P_1$ for time $t$. The second linear interpolation is $\overline{P_1P_2}$ and is the interpolation between $P_1$ and $P_2$ for time $t$.  The final point on the curve will be the interpolation at time $t$ between $\overline{P_0P_1}$ and $\overline{P_1P_2}$, which we can call $\stackrel{\frown}{P_0P_1P_2}$.

@@

The pattern continues for higher texture dimensions and curve orders as well.  Linear texture sampling of textures of dimension $N$ is just the linear interpolation between two linear texture samples of dimension $N-1$.  Using the De Casteljeau algorithm to evaluate a curve of order $N$ just means that you take a linear interpolation between curves of order $N-1$.

@@

It is also possible to get points on curves of higher order by taking multiple texture reads and combining them either using the De Casteljeau algorithm, or by using the Bernstein form equation of Bezier curves.

@@

This pattern continues for trilinear interpolation mapping to a cubic curve, and beyond.  $N$ dimensional linear texture sampling is just the linear interpolation between two linear texture samples of dimension $N-1$.  Bezier curves of order $N$ are just the linear interpolation between two bezier curves of order $N-1$.

@@

Using one dimensional textures, the built in texture interpolation is only able to perform linear interpolation, which allows calculations of only 1st order (linear) Bezier curves (\autoref{fig:decdimensionality}).

Since the De Casteljeau algorithm just interpolates between linear interpolations, we can take more texture samples and combine them in the shader program to get higher order curves.  For an order $N$ Bezier curve, the De Casteljeau algorithm requires $2^{N-1}$ linear interpolations for the linear level of the tree.  You need two pixels to perform a linear interpolation, so it seems that we would need $2^N$ pixels to store a curve of order $N$ in a 1 dimensional textures.  For a cubic curve, that would look like this: $[P_0,P_1,P_1,P_2,P_1,P_2,P_2,P_3]$.

We can do better than that though.  Firstly, there are redundant lerps in the list for lerping between $P_1$ and $P_2$.  As the order of the curve increases, the number of redundancies increase as well.  We don't want to have to pay the cost of pixel storage and texture samples for redundancies, so because we have $N+1$ control points and we only ever sample between  $P_i$ and $P_{i+1}$, we only need to store $N$ lerps in the 1 dimensional texture, which would mean we would need $N*2$ pixels.  A cubic curve would then look like this:  $[P_0,P_1,P_1,P_2,P_2,P_3]$.

After eliminating redundant lerps, we can see that there are redundant data values in our texture.  $P_1$ and $P_2$ each show up twice.  As the order of the curve increases, we would notice the pattern that all control points show up twice except the end points.  We can remove these redundancies and still be able to get the same information out.  The result of removing these redundancies is that our texture only needs to contain each control point once, which means we only need $N+1$ pixels to encode a curve of order $N$.  The cubic curve now looks like this: $[P_0,P_1,P_2,P_3]$.

@@

In the case of a cubic curve with control points $A,B,C,D$, the one dimensional texture will be 4 pixels in size, and three texture reads will need to be done to get the interpolations for $\overline{AB},\overline{BC},\overline{CD}$.  Those will then need to be combined using either the De Casteljeau algorithm, or by using the Bernstein form of a quadratic Bezier curve ($P = A*s^2 + B*st + C*t^2$) to elevate those three points from order 1 to order 3, to get the final point $\stackrel{\frown}{ABCD}$.  Sample code of this example is provided in \autoref{lst:GLSLCubicTexture1D}.

@@

\subsection{Two Dimensional Textures}
Using two dimensional textures allows the texture interpolator to perform bilinear interpolation, which allows calculations of 2nd order (quadratic) Bezier curves (\autoref{fig:decdimensionality}).

This allows us to either calculate a quadratic Bezier curve with a single texture read, or allows us to get the quadratic level values needed to make a higher order curve.

@@

Knowing that we can linearly interpolate between two quadratic curves to create a cubic curve, and that we can combine three quadratic curves to create a quartic curve, the pattern seems to be that we need a texture size of $(2,2*(N-1))$ to encode a curve of order $N$ into a two dimensional texture.

When actually creating that texture, we would once again see redundancies however.  For a cubic curve with control points $P_0,P_1,P_2,P_3$ we would see that the top row of pixels was $P_0,P_1$, then the next row was $P_1,P_2$.  That would be our first quadratic curve.  The third row would start our second quadratic curve which would be $P_1,P_2$ and then the fourth and final row would be $P_2,P_3$.  In this case, $P_1,P_2$ is redundant.  If we increased the order of the curve, the pattern that we would see is that every curve after the first curve would have a redundant row.

Removing that redundant row, it means that we start with a $2$x$2$ texture for a quadratic curve, and then just add a row of pixels for each order above quadratic.  That makes the size of our texture required be only $(2,N)$.

@@

Using three dimensional textures (often referred to as volumetric textures) allows the texture interpolator to evaluate trilinear interpolation, which allows calculations of 3rd order (cubic) Bezier curves (\autoref{fig:decdimensionality}).

This allows us to either calculate a cubic Bezier curve with a single texture read, or allows us to get the cubic level values needed to make a higher order curve.

@@

Increasing texture dimensionality allows you to evaluate the same order curve with fewer texture reads but comes at the cost of using more texture memory.  We evaluated texture dimensions one, two and three, but the pattern continues for higher dimensions as well.

The table below is a comparison of texture dimensionality options.  $M$ is the number of piecewise curves being stored and $N$ is the order of the curve.  To get an idea of how dimensionality and order affects the size of the required textures, take a look at \autoref{fig:pixelcount}.

\begin{tabular}{|l|l|l|l|l|}
\hline
 & \bf{Min. Order} & \bf{\# Samples} & \bf{Dimensions} & \bf{\# Pixels} \\ \hline
\bf{1D} & 1 (Linear) & $N$ & $(M*(N+1))$ & $M*(N+1)*1$ \\ \hline
\bf{2D} & 2 (Quadratic) & $N-1$ & $(2*M,N)$ & $M*(N+0)*2$\\ \hline
\bf{3D} & 3 (Cubic) & $N-2$ & $(2*M,N-1,2)$ & $M*(N-1)*4$\\ \hline
\bf{4D} & 4 (Quartic) & $N-3$ & $(2*M,N-2,2,2)$ & $M*(N-2)*8$\\ \hline
\end{tabular}

@@

      \begin{axis}[xmin=1,xmax=6,ymin=0,xtick={1,...,6},ytick={0,5,10,15,20,25,30},xlabel={Order},ylabel={Pixel Count},legend pos=north west]

@@

    \caption{Pixel count for curve order, with different texture dimensionalities.}
    \label{fig:pixelcount}
  \end{figure}

@@

If you encode Bezier curves into a texture which has multiple color channels - such as the standard four R,G,B,A - you can set up the textures such that once you read in the R,G,B,A, you can combine the channel values together such that you get a fewer number of curves per texture read, but that they are higher order.

For instance, if you had a 2x2 texture, you could encode the control points A,B,C into the red channel, and the control points B,C,D into the green channel.  Then, after you did a texture read, you could do a lerp between the red and the green channel.  Since the red and green channel each contain a quadratic curve, the result would be a cubic curve, defined by the control points A,B,C,D.  However, you've used up two of the channels for a single curve, so are getting fewer curves per texture read, but they are higher order.  The blue and the alpha channel could either be quadratic curves, or could also be encoded such that they could be combined together to provide another cubic curve.

@@

If you you are combining M color channels, each of which are values from curves of order N, the resulting curve value will be order N+M-1.  The intuitive explanation to this, is that each channel has the ability to add one more control point to the curve, regardless of what texture dimensionality you are working with.

@@

When you are in need of a curve with lots of control points, piecewise curves have several benefits over increasing the order of the curve.

The main benefit of piecewise curves are that they allow us to use lower order curves, which allows us to use lower dimensional textures, and have less computation in the shader programs.

@@

Also, if you want a higher order curve than your texture can provide with a single read, forcing you to combine multiple curve points, combining those curve points in the shader program has the nice side benefit of increasing accuracy as well.

>>>>

\begin{abstract}
\small
Modern graphics techniques expose internal parameters to allow re-use and to help separate the technical implementation from artistic usage cases.  A popular choice is to expose parameters as customizable curves and to quantize the curves into textures.  Quantization leads to either lower quality results, or more texture memory being used to accommodate higher sampling frequencies.  The technique presented in this paper leverages the capabilities of GPU texture samplers to allow more efficient storage and evaluation of both integral and rational Bezier curves of any degree, resulting in higher fidelity for similar costs.  Piecewise curves, B-Splines and NURBS are addressed, and there are also limited applications towards vector graphics.
\end{abstract}

@@

This paper shows a third method where:
\begin{itemize}
  \item The control points of the curves are encoded in a texture.
  \item The texture sampler calculates points on the curve before the data reaches the shader program.
  \item It gives accuracy results closer to that of shader constant curves, while having less calculation overhead.
  \item The technique can support both integral and rational Bezier curves of any degree and can also be used for piecewise curves.
  \item The curve type must be decided on in advance, like when using the shader constant method.
  \item There are limited applications towards vector graphics.
\end{itemize}

@@

The core of this technique is that the $N$-dimensional linear texture interpolation capabilities on the GPU can be mathematically equivalent to De Casteljeau's algorithm for evaluating Bezier curves.  We can use linear texture sampling on textures of increasing dimensionalities to have the texture sampler calculate points on curves of increasing degree.

@@

One dimensional linear texture sampling is just the linear interpolation between two data points.  This is equivalent to the De Casteljeau algorithm when evaluating a linear curve of degree 1.  In both cases, it is just linear interpolation between two values.

@@

The De Casteljeau algorithm for evaluating a quadratic curve (degree 2) is also just a linear interpolation between two linear interpolations.  If the control points of the quadratic curve are $P_0,P_1,P_2$, let's call the first interpolation $\overline{P_0P_1}$ which is the interpolation between $P_0$ and $P_1$ for time $t$. The second linear interpolation is $\overline{P_1P_2}$ and is the interpolation between $P_1$ and $P_2$ for time $t$.  The final point on the curve will be the interpolation at time $t$ between $\overline{P_0P_1}$ and $\overline{P_1P_2}$, which we can call $\stackrel{\frown}{P_0P_1P_2}$.

@@

The pattern continues for higher texture dimensions and curve degrees as well.  Linear texture sampling of textures of dimension $N$ is just the linear interpolation between two linear texture samples of dimension $N-1$.  Using the De Casteljeau algorithm to evaluate a curve of degree $N$ just means that you take a linear interpolation between curves of degree $N-1$.

@@

It is also possible to get points on curves of higher degree by taking multiple texture reads and combining them either using the De Casteljeau algorithm, or by using the Bernstein form equation of Bezier curves.

@@

This pattern continues for trilinear interpolation mapping to a cubic curve, and beyond.  $N$ dimensional linear texture sampling is just the linear interpolation between two linear texture samples of dimension $N-1$.  Bezier curves of degree $N$ are just the linear interpolation between two Bezier curves of degree $N-1$.

@@

Using one dimensional textures, the built in texture interpolation is only able to perform linear interpolation, which allows calculations of only 1st degree (linear) Bezier curves (\autoref{fig:decdimensionality}).

Since the De Casteljeau algorithm just interpolates between linear interpolations, we can take more texture samples and combine them in the shader program to get higher degree curves.  For a degree $N$ Bezier curve, the De Casteljeau algorithm requires $2^{N-1}$ linear interpolations for the linear level of the tree.  You need two pixels to perform a linear interpolation, so it seems that we would need $2^N$ pixels to store a curve of degree $N$ in a 1 dimensional textures.  For a cubic curve, that would look like this: $[P_0,P_1,P_1,P_2,P_1,P_2,P_2,P_3]$.

We can do better than that though.  Firstly, there are redundant lerps in the list for lerping between $P_1$ and $P_2$.  As the degree of the curve increases, the number of redundancies increase as well.  We don't want to have to pay the cost of pixel storage and texture samples for redundancies, so because we have $N+1$ control points and we only ever sample between  $P_i$ and $P_{i+1}$, we only need to store $N$ lerps in the 1 dimensional texture, which would mean we would need $N*2$ pixels.  A cubic curve would then look like this:  $[P_0,P_1,P_1,P_2,P_2,P_3]$.

After eliminating redundant lerps, we can see that there are redundant data values in our texture.  $P_1$ and $P_2$ each show up twice.  As the degree of the curve increases, we would notice the pattern that all control points show up twice except the end points.  We can remove these redundancies and still be able to get the same information out.  The result of removing these redundancies is that our texture only needs to contain each control point once, which means we only need $N+1$ pixels to encode a curve of degree $N$.  The cubic curve now looks like this: $[P_0,P_1,P_2,P_3]$.

@@

In the case of a cubic curve with control points $A,B,C,D$, the one dimensional texture will be 4 pixels in size, and three texture reads will need to be done to get the interpolations for $\overline{AB},\overline{BC},\overline{CD}$.  Those will then need to be combined using either the De Casteljeau algorithm, or by using the Bernstein form of a quadratic Bezier curve ($P = A*s^2 + B*st + C*t^2$) to elevate those three points from degree 1 to degree 3, to get the final point $\stackrel{\frown}{ABCD}$.  Sample code of this example is provided in \autoref{lst:GLSLCubicTexture1D}.

@@

\subsection{Two Dimensional Textures}
Using two dimensional textures allows the texture interpolator to perform bilinear interpolation, which allows calculations of 2nd degree (quadratic) Bezier curves (\autoref{fig:decdimensionality}).

This allows us to either calculate a quadratic Bezier curve with a single texture read, or allows us to get the quadratic level values needed to make a higher degree curve.

@@

Knowing that we can linearly interpolate between two quadratic curves to create a cubic curve, and that we can combine three quadratic curves to create a quartic curve, the pattern seems to be that we need a texture size of $(2,2*(N-1))$ to encode a curve of degree $N$ into a two dimensional texture.

When actually creating that texture, we would once again see redundancies however.  For a cubic curve with control points $P_0,P_1,P_2,P_3$ we would see that the top row of pixels was $P_0,P_1$, then the next row was $P_1,P_2$.  That would be our first quadratic curve.  The third row would start our second quadratic curve which would be $P_1,P_2$ and then the fourth and final row would be $P_2,P_3$.  In this case, $P_1,P_2$ is redundant.  If we increased the degree of the curve, the pattern that we would see is that every curve after the first curve would have a redundant row.

Removing that redundant row, it means that we start with a $2$x$2$ texture for a quadratic curve, and then just add a row of pixels for each degree above quadratic.  That makes the size of our texture required be only $(2,N)$.

@@

Using three dimensional textures (often referred to as volumetric textures) allows the texture interpolator to evaluate trilinear interpolation, which allows calculations of 3rd degree (cubic) Bezier curves (\autoref{fig:decdimensionality}).

This allows us to either calculate a cubic Bezier curve with a single texture read, or allows us to get the cubic level values needed to make a higher degree curve.

@@

Increasing texture dimensionality allows you to evaluate the same degree curve with fewer texture reads but comes at the cost of using more texture memory.  We evaluated texture dimensions one, two and three, but the pattern continues for higher dimensions as well.

The table below is a comparison of texture dimensionality options.  $M$ is the number of piecewise curves being stored and $N$ is the degree of the curve.  To get an idea of how dimensionality and degree affects the size of the required textures, take a look at \autoref{fig:pixelcount}.

\begin{tabular}{|l|l|l|l|l|}
\hline
 & \bf{Min. Degree} & \bf{\# Samples} & \bf{Dimensions} & \bf{\# Pixels} \\ \hline
\bf{1D} & 1 (Linear) & $N$ & $(M*(N+1))$ & $M*(N+1)*1$ \\ \hline
\bf{2D} & 2 (Quadratic) & $N-1$ & $(2*M,N)$ & $M*(N+0)*2$\\ \hline
\bf{3D} & 3 (Cubic) & $N-2$ & $(2*M,N-1,2)$ & $M*(N-1)*4$\\ \hline
\bf{4D} & 4 (Quartic) & $N-3$ & $(2*M,N-2,2,2)$ & $M*(N-2)*8$\\ \hline
\end{tabular}

@@

      \begin{axis}[xmin=1,xmax=6,ymin=0,xtick={1,...,6},ytick={0,5,10,15,20,25,30},xlabel={Degree},ylabel={Pixel Count},legend pos=north west]

@@

    \caption{Pixel count for curve degree, with different texture dimensionalities.}
    \label{fig:pixelcount}
  \end{figure}

@@

If you encode Bezier curves into a texture which has multiple color channels - such as the standard four R,G,B,A - you can set up the textures such that once you read in the R,G,B,A, you can combine the channel values together such that you get a fewer number of curves per texture read, but that they are higher degree.

For instance, if you had a 2x2 texture, you could encode the control points A,B,C into the red channel, and the control points B,C,D into the green channel.  Then, after you did a texture read, you could do a lerp between the red and the green channel.  Since the red and green channel each contain a quadratic curve, the result would be a cubic curve, defined by the control points A,B,C,D.  However, you've used up two of the channels for a single curve, so are getting fewer curves per texture read, but they are higher degree.  The blue and the alpha channel could either be quadratic curves, or could also be encoded such that they could be combined together to provide another cubic curve.

@@

If you you are combining M color channels, each of which are values from curves of degree N, the resulting curve value will be degree N+M-1.  The intuitive explanation to this, is that each channel has the ability to add one more control point to the curve, regardless of what texture dimensionality you are working with.

@@

When you are in need of a curve with lots of control points, piecewise curves have several benefits over increasing the degree of the curve.

The main benefit of piecewise curves are that they allow us to use lower degree curves, which allows us to use lower dimensional textures, and have less computation in the shader programs.

@@

Also, if you want a higher degree curve than your texture can provide with a single read, forcing you to combine multiple curve points, combining those curve points in the shader program has the nice side benefit of increasing accuracy as well.
